---
title: "Digital Data Capture & Analysis Lab Book"
author: "Edward Mason"
output: 
  html_document: 
    toc: true
    toc_depth: 6
    toc_float: true
    theme: paper
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r load packages, echo = F, warning = F, message = F}
library(tidyverse)
library(plotly)
library(ggpubr)
library(ggpmisc)
library(cowplot)
library(lubridate)
library(colourpicker)
library(gganimate)
library(thermocouple)
library(Metrics)

library(sf)
library(photosearcher)
library(textclean)
library(tidytext)
library(igraph, warn.conflicts = FALSE)

library(tidytransit)
```

## Weeks 1 and 2: Creating a 360 Virtual Tour of Campus

This week we have been using [ThingLink](https://thinglink.com) and 360
photos taken with GoPro/Insta360 cameras to create 360 Virtual Tours.

Our tour is of the Y3 building on the University campus. We started in
the basement, going "up" the stairs and out of building, moving past the
front of the building and back in via another door. We then went back
into the basement via the lift, and then finishing in one of the seminar
rooms. We used a GoPro Max camera on a tripod, and connected it via
Bluetooth to an iPhone with the GoPro Quik app on for remote control of
the shutter.

I then uploaded the footage to ThingLink to create the tour. The
platform was fairly intuitive to use, but it was a slow and repetitive
task to go through each image in turn, changing settings and adding
"tags" to enable moving between each image. Furthermore, there were some
shots with people in, which were blurred out for privacy reasons.
Viewing it with a VR headset was impressive, despite the low resolution
of the images.

A virtual tour such as this would be useful for familiarising people
with (for example) a campus, office space, or a new development.
However, ethical concerns such as photographing of bystanders, as well
as taking photos of sensitive sites should be considered.

Below is the final tour on ThingLink.

```{r thinglink tour, echo = F}
knitr::include_url("https://www.thinglink.com/view/scene/2073448908817695397")
```

## Weeks 3 and 4: Photogrammetry (Structure for Motion modelling)

```{r sfm model, echo = F}
knitr::include_url("https://sketchfab.com/models/6c712d4d24a74af5b12fbec2aef62aae/embed")
```

The model above is of an IKEA Marius stool in my living room. Overall, I
would say this model is quite poor, as it does not adequately represent
the stool.

Firstly, the mesh of the model is quite poor, leading to holes and
floating geometry, as ghost points are formed. There was also erroneous
mesh points formed from carpet and wall shots. As well as this, the
texture of the model is very poorly captured, with none of the black
colour of the stool in the model.

Both of these issues could be due to the lack of a plain background
(making it more difficult for the software to distinguish what the focus
is), a poor photo distribution (i.e. the photos did not provide
sufficient coverage of the stool), or possibly poor camera quality. The
other point to consider is the software used - an open-source package
called [Regard3D](https://regard3d.org). Alternative methods of
processing could be considered, such as modifying parameters, however
considering the aforementioned background concerns, these would probably
be not as consequential in approving the final model output.

## Weeks 5 and 7: Enivronmental Sensing

```{r environmental sensing - load data, echo = F, warning = F, message = F}
df_cal <- read_csv("Practical-Week7/NTC_cal_example.csv")
```

Turning to environmental sensors, we looked into how open-source
hardware and software and how their use has lowered the barriers to
sensing. This opens up possibilities to non-profits, for example, to use
sensors to back up their concerns with data. In our case, we used a
temperature sensor, composed of a thermistor attached to an Arduino
micro-computer (see Figure 1). The resistance across the thermistor
changes with temperature, and therefore if this resistance is measured a
temperature can be calculated from it.

![Figure 1: Temperature Sensor
Schematic](labbook_images/clipboard-3779242095.png)

\

The first part of the process involved calibration. Using four water
baths at different temperatures, the resistance read by the
Arduino-based sensor was recorded along with the temperature from a
reference probe, with ten readings taken from each bath. The results are
shown in Figure 2:

```{r resistance-temp graph, echo = F, fig.width = 10, fig.cap = "Figure 2: A graph to show Resistance against Temperature as measured by an Arduino-based temperature sensor"}
ggplot(data = df_cal, mapping = aes(x = Resistance, y = Ref_temp)) + 
  geom_point(size=3.5, shape=1, colour="black") + 
  ylab("Water temperature (\u00b0C)") +
  xlab("Resistance (\u2126)")
```

\

We can see a clear pattern of an increase of resistance as temperature
decreases - this means that the thermistor used has a negative
temperature coefficient (an NTC thermistor).

Next, we will fit a calibration model to the data. The code below
creates a linear and non-linear (polynomial) model from the data.

```{r models}
linear_model <- lm(data = df_cal, 
                   formula = Ref_temp ~ Resistance)

poly_model <- lm(data = df_cal, 
                 formula = Ref_temp ~ poly(
                   x = Resistance, degree = 2, raw = T))
```

The summaries are shown in the table below:

| Model | Residual Std. Error | Multiple R-Squared | F-Statistic and p-value |
|:--:|:--:|:--:|:--:|
| Linear Model | 3.784 (28 *df*) | 0.9434 | 466.4, p \< 2.2e-16 |
| Polynomial Model | 0.2554 (27 *df*) | 0.9998 | 5.424 × 10^4^ , p \< 2.2e-16 |

: Table 1: Calibration Model Summaries

The summary statistics from this model shows that the polynomial model
is a better fit to the data, with a smaller residual standard error, and
a larger R-Squared value and F-Statistic.

Below are the models plotted against the data:

```{r models and data graphs, echo = F, warning = F, message = F}
ggplot()
```

### Todo

1.  ~~Give this section of your lab book a title – e.g "using
    environmental sensors".~~

2.  ~~Briefly explain how you collected the data (i.e. the methods you
    used). Here you can include images, schematics and text. Be sure to
    highlight the instruments used.~~

3.  ~~For the calibration generate a scatter plot of resistance vs
    observed temperature. Ensure you include a figure caption and some
    plain text with a brief description of the pattern(s) observed.~~

4.  ~~Now fit a linear model and a non-linear model. Extract information
    on the model fit and present these in a table with a short
    interpretation of the results (which model appear to have the better
    fit)?~~

5.  Apply both models to the data and calculate the mean and SD for each
    calibration data point. Present this as a table, what do you notice
    about the data?

6.  Calculate the sensor linearity and accuracy (RMSE) based on the
    model you think is best. Present this data as a table or integrated
    into a short paragraph that describes the results. Would you be
    happy to use this sensor for another research project? How does it
    compare to commercial sensor specifications?

7.  For the ADC experiment plot the data as boxplot (8-bit, 10-bit and
    12-bit). What do you notice? Now plot the data as a dot plot – what
    does this tell us about the difference in the step size (i.e.
    resolution in °C)?

8.  Calculate the mean and SD for each resolution along with the mean
    and SD of the reference sensor (10 data points required). What do
    you notice?

9.  ~~End your notebook with some short reflections on using the Arduino
    platform – what worked well, what didn’t and what you’d do
    differently if you were using the platform for another project.~~

The Arduino platform is a well-developed "open-hardware" ecosystem with
supporting software and documentation. However, as with all
"open-source" systems, there are some pitfalls. Firstly, the system is
very much DIY - you are provided with the components, but as the user
you must then assemble the parts, both physically (the hardware) and in
terms of programming the assembled system to do what you would like it
to. This requires, in the case of the Arduino ecosystem, a knowledge of
C/C++ languages. Furthermore, if anything goes wrong, there is no
dedicated customer support, but only documentation and support forums.
That being said, if these obstacles can be overcome, the Arduino
platform can scale rapidly and relatively cheaply compared to
alternative proprietary sensing systems, making it a viable option for
more widespread sensing networks, or for smaller companies and/or
non-profits who cannot afford the more expensive alternatives.

## Week 6: APIs and AI

### Todo

1.  An appropriate title

2.  A map of Sutton Park showing the locations of all of the photos,
    coloured by the cluster that photo belongs to. Make sure to give
    your clusters informative names based on the top keywords

3.  A bar plot of the results of the sentiment analysis

4.  A reflection on the use of Flickr as a data source to understand
    people’s experiences of green spaces

5.  A reflection on the spatial patterns in the clusters. Are there
    areas with more photos than others?

6.  A reflection on whether some clusters are more consistently
    positive/negative than others. Why might this be?

7.  Suggestion for other data sources or methods you might you consider
    to understand what people value about green spaces (staying within
    the scope of this practical – APIs and AI)

## Final Project: Using GTFS Feeds to Visualise Transport Data from the West Midlands

**Project Aim**: Using Transport for West Midlands' GTFS feeds and the
tidytransit R package [@poletti2026], this project shows how service
frequency and transport mode varies across the West Midlands Combined
Authority area.

### Todo

1.  Briefly explain what you’re trying to achieve with the project

2.  Explain the choices you made on the techniques used

3.  Reflect on any ethical implications

4.  Critically evaluate the effectiveness of your project

## References
