---
title: "Digital Data Capture & Analysis Lab Book"
author: "Edward Mason"
output: 
  html_document:
    toc: true
    toc_depth: 6
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
---

## Weeks 1 and 2: Creating a 360 Virtual Tour of Campus

This week we have been using [ThingLink](https://thinglink.com) and 360
photos taken with GoPro/Insta360 cameras to create 360 Virtual Tours.

Our tour is of the Y3 building on the University campus. We started in
the basement, going "up" the stairs and out of building, moving past the
front of the building and back in via another door. We then went back
into the basement via the lift, and then finishing in one of the seminar
rooms. We used a GoPro Max camera on a tripod, and connected it via
Bluetooth to an iPhone with the GoPro Quik app on for remote control of
the shutter.

I then uploaded the footage to ThingLink to create the tour. The
platform was fairly intuitive to use, but it was a slow and repetitive
task to go through each image in turn, changing settings and adding
"tags" to enable moving between each image. Furthermore, there were some
shots with people in, which were blurred out for privacy reasons.
Viewing it with a VR headset was impressive, despite the low resolution
of the images.

A virtual tour such as this would be useful for familiarising people
with (for example) a campus, office space, or a new development.
However, ethical concerns such as photographing of bystanders, as well
as taking photos of sensitive sites should be considered.

Below is the final tour on ThingLink.

```{r echo=FALSE}
knitr::include_url("https://www.thinglink.com/view/scene/2073448908817695397")
```

## Weeks 3 and 4: Photogrammetry (Structure for Motion modelling)

```{r echo=FALSE}
knitr::include_url("https://sketchfab.com/models/6c712d4d24a74af5b12fbec2aef62aae/embed")
```

The model above is of an IKEA Marius stool in my living room. Overall, I would say this model is quite poor, as it does not adequately represent the stool.

Firstly, the mesh of the model is quite poor, leading to holes and floating geometry, as ghost points are formed. There was also erroneous mesh points formed from carpet and wall shots. As well as this, the texture of the model is very poorly captured, with none of the black colour of the stool in the model.

Both of these issues could be due to the lack of a plain background (making it more difficult for the software to distinguish what the focus is), a poor photo distribution (i.e. the photos did not provide sufficient coverage of the stool), or possibly poor camera quality. The other point to consider is the software used - an open-source package called [Regard3D](https://regard3d.org). Alternative methods of processing could be considered, such as modifying parameters, however considering the aforementioned background concerns, these would probably be not as consequential in approving the final model output.

## Weeks 5 and 7: Enivronmental Sensing
_Coming soon!_

## Week 6: APIs and AI
_Coming soon!_

## Final Project
_Coming soon!_