---
title: "Digital Data Capture & Analysis Lab Book"
author: "Edward Mason"
output: 
  html_document: 
    toc: true
    toc_depth: 6
    toc_float: true
    theme: paper
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r load packages, echo = F, warning = F, message = F}
library(tidyverse)
library(plotly)
library(ggpubr)
library(ggpmisc)
library(cowplot)
library(lubridate)
library(colourpicker)
library(gganimate)
library(thermocouple)
library(Metrics)

library(sf)
library(photosearcher)
library(textclean)
library(tidytext)
library(igraph, warn.conflicts = FALSE)

library(tidytransit)
```

## Weeks 1 and 2: Creating a 360 Virtual Tour of Campus

This week we have been using [ThingLink](https://thinglink.com) and 360
photos taken with GoPro/Insta360 cameras to create 360 Virtual Tours.

Our tour is of the Y3 building on the University campus. We started in
the basement, going "up" the stairs and out of building, moving past the
front of the building and back in via another door. We then went back
into the basement via the lift, and then finishing in one of the seminar
rooms. We used a GoPro Max camera on a tripod, and connected it via
Bluetooth to an iPhone with the GoPro Quik app on for remote control of
the shutter.

I then uploaded the footage to ThingLink to create the tour. The
platform was fairly intuitive to use, but it was a slow and repetitive
task to go through each image in turn, changing settings and adding
"tags" to enable moving between each image. Furthermore, there were some
shots with people in, which were blurred out for privacy reasons.
Viewing it with a VR headset was impressive, despite the low resolution
of the images.

A virtual tour such as this would be useful for familiarising people
with (for example) a campus, office space, or a new development.
However, ethical concerns such as photographing of bystanders, as well
as taking photos of sensitive sites should be considered.

Below is the final tour on ThingLink.

```{r thinglink tour, echo = F, fig.width = 10, fig.cap = "Figure 1: A virtual tour of University of Birmingham Y3 Building"}
knitr::include_url("https://www.thinglink.com/view/scene/2073448908817695397")
```

## Weeks 3 and 4: Photogrammetry (Structure for Motion modelling)

```{r sfm model, echo = F, fig.width = 10, fig.cap = "Figure 2: A model of an IKEA Marius stool"}
knitr::include_url("https://sketchfab.com/models/6c712d4d24a74af5b12fbec2aef62aae/embed")
```

The model above is of an IKEA Marius stool in my living room. Overall, I
would say this model is quite poor, as it does not adequately represent
the stool.

Firstly, the mesh of the model is quite poor, leading to holes and
floating geometry, as ghost points are formed. There was also erroneous
mesh points formed from carpet and wall shots. As well as this, the
texture of the model is very poorly captured, with none of the black
colour of the stool in the model.

Both of these issues could be due to the lack of a plain background
(making it more difficult for the software to distinguish what the focus
is), a poor photo distribution (i.e. the photos did not provide
sufficient coverage of the stool), or possibly poor camera quality. The
other point to consider is the software used - an open-source package
called [Regard3D](https://regard3d.org). Alternative methods of
processing could be considered, such as modifying parameters, however
considering the aforementioned background concerns, these would probably
be not as consequential in approving the final model output.

## Weeks 5 and 7: Enivronmental Sensing

```{r environmental sensing - load data, echo = F, warning = F, message = F}
df_cal <- read_csv("Practical-Week7/NTC_cal_example.csv")
df_adc_water <- read_csv("Practical-Week7/ADC_water_bath_example.csv")
df_adc_air <- read_csv("Practical-Week7/ADC_air_logging_example.csv")
df_adc_refT <- read_csv("Practical-Week7/ADC_ref_temp.csv")

# Correct time formatting error in df_adc_refT data
df_adc_refT$Time <- parse_time(df_adc_refT$Time, "%H.%M.%S")

# Restructure the data
## ADC water bath
df_adc_water_L <- df_adc_water |>
  pivot_longer(!c(Time),
               names_to = "ADC_Bits", values_to = "Resistance")

df_adc_water_L <- df_adc_water_L |>
  mutate(min=minute(Time))

## ADC air
df_adc_air_L <- df_adc_water |>
  pivot_longer(!c(Time),
               names_to = "ADC_Bits", values_to = "Resistance")

df_adc_air_L <- df_adc_air_L |>
  mutate(min=minute(Time))
```

Turning to environmental sensors, we looked into how open-source
hardware and software and how their use has lowered the barriers to
sensing. This opens up possibilities to non-profits, for example, to use
sensors to back up their concerns with data. In our case, we used a
temperature sensor, composed of a thermistor attached to an Arduino
micro-computer (see Figure 1). The resistance across the thermistor
changes with temperature, and therefore if this resistance is measured a
temperature can be calculated from it.

![Figure 3: Temperature Sensor
Schematic](labbook_images/clipboard-3779242095.png)

\

### Calibration

The first part of the process involved calibration. Using four water
baths at different temperatures, the resistance read by the
Arduino-based sensor was recorded along with the temperature from a
reference probe, with ten readings taken from each bath. The results are
shown in Figure 2:

```{r resistance-temp graph, echo = F, fig.width = 10, fig.cap = "Figure 4: A graph to show Resistance against Temperature as measured by an Arduino-based temperature sensor"}
ggplot(data = df_cal, mapping = aes(x = Resistance, y = Ref_temp)) + 
  geom_point(size=3.5, shape=1, colour="black") + 
  ylab("Water temperature (\u00b0C)") +
  xlab("Resistance (\u2126)")
```

\

We can see a clear pattern of an increase of resistance as temperature
decreases - this means that the thermistor used has a negative
temperature coefficient (an NTC thermistor).

Next, we will fit a calibration model to the data. The code below
creates a linear and non-linear (polynomial) model from the data.

```{r models}
linear_model <- lm(data = df_cal, 
                   formula = Ref_temp ~ Resistance)

poly_model <- lm(data = df_cal, 
                 formula = Ref_temp ~ poly(
                   x = Resistance, degree = 2, raw = T))
```

The summaries are shown in the table below:

| Model | Residual Std. Error | Multiple R-Squared | F-Statistic and p-value |
|:--:|:--:|:--:|:--:|
| Linear Model | 3.784 (28 *df*) | 0.9434 | 466.4, p \< 2.2e-16 |
| Polynomial Model | 0.2554 (27 *df*) | 0.9998 | 5.424 × 10^4^ , p \< 2.2e-16 |

: Table 1: Calibration Model Summaries

The summary statistics from this model shows that the polynomial model
is a better fit to the data, with a smaller residual standard error, and
a larger R-Squared value and F-Statistic.

Below are the models plotted against the data:

```{r linear model and data graph, echo = F, warning = F, message = F, fig.width = 10, fig.cap = "Figure 5: A graph to show Resistance against Temperature with a linear model"}
ggplot(df_cal, aes(y = Ref_temp, x = Resistance)) +
  geom_point() +
  geom_smooth(method = lm)
```

```{r polynomial model and data graph, echo = F, warning = F, message = F, fig.width = 10, fig.cap = "Figure 6: A graph to show Resistance against Temperature with a non-linear model"}
ggplot(df_cal, aes(y = Ref_temp, x = Resistance)) +
  geom_point() +
  geom_smooth(method = lm, formula = y ~ poly(x, 2, raw = T))
```

The mean and standard deviation for each model is shown below.

```{r beta and log models; fitted values into a table, echo = F, warning = F, message = F}
df_cal$Log_res <- log(df_cal$Resistance)
log_linear_model<-lm(data = df_cal, formula = Ref_temp ~ Log_res)
df_cal$Temp_beta <- ThermistorTemperature(R0 = 10000, T0 = 25, R = df_cal$Resistance, betaTH = 3950)
df_cal$Temp_linear <- linear_model$fitted.values
df_cal$Temp_poly <- poly_model$fitted.values
df_cal$Temp_loglin <- log_linear_model$fitted.values

tib_cal <- tibble(Ref_temp = df_cal$Ref_temp, Temp_beta = df_cal$Temp_beta, Temp_linear = df_cal$Temp_linear, Temp_poly = df_cal$Temp_poly, Temp_loglin = df_cal$Temp_loglin)

tib_cal_summary <- tib_cal |>
  group_by(Ref_temp) |>
  summarise(
    Mean_temp_beta = mean(Temp_beta),
    Sd_temp_beta = sd(Temp_beta),
    Mean_temp_linear = mean(Temp_linear),
    Sd_temp_linear = sd(Temp_linear),
    Mean_temp_poly= mean(Temp_poly),
    Sd_temp_poly = sd(Temp_poly),
    Mean_temp_loglin = mean(Temp_loglin),
    Sd_temp_loglin = sd(Temp_loglin)
  )

tib_cal_summary <- tib_cal_summary |> mutate(
  across(where(is.numeric), round, 3)
)

tib_cal_summary <- tib_cal_summary |>
  unite(Summ_temp_beta, Mean_temp_beta:Sd_temp_beta, sep = "±", na.rm = T) |>
  unite(Summ_temp_linear, Mean_temp_linear:Sd_temp_linear, sep = "±", na.rm = T) |>
  unite(Summ_temp_poly, Mean_temp_poly:Sd_temp_poly, sep = "±", na.rm = T) |>
  unite(Summ_temp_loglin, Mean_temp_loglin:Sd_temp_loglin, sep = "±", na.rm = T)

knitr::kable(tib_cal_summary, col.names = c("Ref Temp (°C)", "Beta Model (Mean ± SD °C)", "Linear Model (Mean ± SD °C)", "Poly Model (Mean ± SD °C)", "Log Linear Model (Mean ± SD °C)"), caption = "Table 2: Summary Statistics for each Reference Temperature and Model")
```

```{r calculate values, echo = T, warning = F, message = F, results = "hide", fig.width = 10, fig.cap = "Figure 7: Models Comparison"}
df_cal_long <- df_cal |> pivot_longer(!c(Cal_medium, Resistance,
                                       Voltage, Ref_temp, 
                                       Log_res),
                                    names_to = "cal_model", 
                                    values_to = "cal_temp")
beta_model <- ThermistorTemperature(R0 = 10000, T0 = 25, R = df_cal$Resistance, betaTH = 3950)

ggplot(data = df_cal_long,
       mapping = aes(x=Ref_temp, y = cal_temp,
       colour = cal_model))+geom_point(size=2)+
  facet_wrap(~cal_model,scales = "fixed")+
  theme_pubclean()+
  geom_abline(slope = 1)+
  geom_smooth(se = F,method = "lm",linetype = "dashed")+  
  stat_regline_equation(label.x=5, label.y=40)

summary(beta_model)
summary(linear_model)
summary(poly_model)
summary(log_linear_model)

rmse(df_cal$Ref_temp, df_cal$Temp_beta)
rmse(df_cal$Ref_temp, df_cal$Temp_linear)
rmse(df_cal$Ref_temp, df_cal$Temp_poly)
rmse(df_cal$Ref_temp, df_cal$Temp_loglin)
```

| **Variable** | Beta Model | Linear Model | Poly Model | Log Linear Model |
|--------------|------------|--------------|------------|------------------|
| R^2^         | N/A        | 0.9434       | 0.9998     | 0.9983           |
| RMSE         | 2.253915   | 3.655814     | 0.2423138  | 0.6319984        |
| Slope        | 0.86       | 0.94         | 1          | 1                |

: Table 3: Statistical Variables for Each Model

Table 3 shows that the sensor is quite accurate in calculating
temperature, especially when paired with a polynomial model. Whilst this
may not be as accurate as a commercial sensor, it gets incredibly close
and for most projects this should be fine.

### Analogue-Digital Conversion (ADC)

```{r adc boxplots data, echo = F, warning = F, message = F}
df_adc_water_L$Temp <- predict(object = poly_model, df_adc_water_L)

df_adc_air_L$Temp <- predict(object =poly_model, df_adc_air_L)


df_adc_air_L$Temp_beta <-ThermistorTemperature(R0 = 10000, T0 = 25, R = df_adc_air_L$Resistance, betaTH = 3950)

df_adc_water_L$Temp_beta <-ThermistorTemperature(R0 = 10000, T0 = 25, R = df_adc_water_L$Resistance, betaTH = 3950)

df_adc_ref_water <- df_adc_refT |> 
  filter(Medium == "Water") |> 
  left_join(df_adc_water_L, by = "Time")

df_adc_ref_air <- df_adc_refT |> 
  filter(Medium == "Air") |> 
  left_join(df_adc_air_L, by = "Time")
```

```{r adc boxplot one, echo = F, warning = F, message = F, fig.width = 10, fig.cap = "Figure 8: Plot of ADC (using Water reference)"}
df_adc_ref_water |> 
  filter(!is.na(ADC_Bits)) |>
  ggplot(aes(x = ADC_Bits, y = Temp_beta)) + 
  geom_boxplot() + 
  geom_point()
```

```{r adc boxplot two, echo = F, warning = F, message = F, fig.width = 10, fig.cap = "Figure 9: Plot of ADC (Using Water reference, without box)"}
df_adc_ref_water |> 
  filter(!is.na(ADC_Bits)) |>
  ggplot(aes(x = ADC_Bits, y = Temp_beta)) + 
  geom_dotplot(binaxis = "y")
```

```{r adc summary table, echo = F, warning = F, message = F}
tib_adc_ref_water <- tibble(df_adc_ref_water)

tib_adc_summary <- tib_adc_ref_water |>
  group_by(ADC_Bits) |>
  summarise(
    Mean = mean(Temp_ref),
    Sd = sd(Temp_ref, na.rm = T)
  )

tib_adc_summary <- tib_adc_summary |> mutate(
  across(where(is.numeric), round, 3)
)

tib_adc_summary <- tib_adc_summary |>
  unite(Summ_temp_ref, Mean:Sd, sep = "±", na.rm = T)

knitr::kable(tib_adc_summary, caption = "Table 5: Summary Statistics for each Reference Temperature and Model")
```

### Todo

1.  ~~Give this section of your lab book a title – e.g "using
    environmental sensors".~~

2.  ~~Briefly explain how you collected the data (i.e. the methods you
    used). Here you can include images, schematics and text. Be sure to
    highlight the instruments used.~~

3.  ~~For the calibration generate a scatter plot of resistance vs
    observed temperature. Ensure you include a figure caption and some
    plain text with a brief description of the pattern(s) observed.~~

4.  ~~Now fit a linear model and a non-linear model. Extract information
    on the model fit and present these in a table with a short
    interpretation of the results (which model appear to have the better
    fit)?~~

5.  ~~Apply both models to the data and calculate the mean and SD for
    each calibration data point. Present this as a table, what do you
    notice about the data?~~

6.  ~~Calculate the sensor linearity and accuracy (RMSE) based on the
    model you think is best. Present this data as a table or integrated
    into a short paragraph that describes the results. Would you be
    happy to use this sensor for another research project? How does it
    compare to commercial sensor specifications?~~

7.  For the ADC experiment plot the data as boxplot (8-bit, 10-bit and
    12-bit). What do you notice? Now plot the data as a dot plot – what
    does this tell us about the difference in the step size (i.e.
    resolution in °C)?

8.  Calculate the mean and SD for each resolution along with the mean
    and SD of the reference sensor (10 data points required). What do
    you notice?

9.  ~~End your notebook with some short reflections on using the Arduino
    platform – what worked well, what didn’t and what you’d do
    differently if you were using the platform for another project.~~

The Arduino platform is a well-developed "open-hardware" ecosystem with
supporting software and documentation. However, as with all
"open-source" systems, there are some pitfalls. Firstly, the system is
very much DIY - you are provided with the components, but as the user
you must then assemble the parts, both physically (the hardware) and in
terms of programming the assembled system to do what you would like it
to. This requires, in the case of the Arduino ecosystem, a knowledge of
C/C++ languages. Furthermore, if anything goes wrong, there is no
dedicated customer support, but only documentation and support forums.
That being said, if these obstacles can be overcome, the Arduino
platform can scale rapidly and relatively cheaply compared to
alternative proprietary sensing systems, making it a viable option for
more widespread sensing networks, or for smaller companies and/or
non-profits who cannot afford the more expensive alternatives.

## Week 6: APIs and AI - Using Flickr's API to map sentiments and topics spatially in Sutton Park, West Midlands

Looking at how to use APIs and AI, this next section shows how the use
of [Flickr](https://flickr.com)'s API can show in a visual way how
people experience green spaces, with very precise spatial data.

```{r apis and ai setup, echo = F, message = F, warning = F, results = "hide"}
boundary <- st_read("Practical-Week6/Sutton Park Boundary_region.shp")

# note - I saved the data from the Flickr API to a csv file in order to save rendering time, as fetching all the data from the API every time takes a while!
sutton_photos <- read_csv("Practical-Week6/sutton_photos.csv", col_names = T, na = "NA")

sutton_text <- sutton_photos %>%
  mutate(
    text = paste(title, description, tags),
    text = replace_url(text),
    text = str_squish(text)
  ) %>%
  filter(!is.na(text), text != "")

data("stop_words")

tidy_tokens <- sutton_text %>%
  select(url_l, longitude, latitude, title, description, tags, text) %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words, by = "word") %>%
  filter(str_detect(word, "^[a-z][a-z'-]*$")) %>%
  count(url_l, word, sort = FALSE)

photo_word <- tidy_tokens %>%
  mutate(n = 1) %>%
  pivot_wider(names_from = word, values_from = n, values_fill = 0)

word_counts <- tidy_tokens %>% count(word, name = "photo_count")

keep_words <- word_counts %>% filter(photo_count >= 5) %>% pull(word)

photo_word_mat <- tidy_tokens %>%
  filter(word %in% keep_words) %>%
  mutate(n = 1) %>%
  pivot_wider(names_from = word, values_from = n, values_fill = 0) %>%
  select(-url_l) %>%
  as.matrix()

keyword_cooccurrance <- t(photo_word_mat) %*% photo_word_mat

graph <- graph_from_adjacency_matrix(keyword_cooccurrance,
                         weighted = TRUE,
                         mode="undirected",
                         diag=FALSE)

clusters <- cluster_walktrap(graph)

keyword_importance <- map_df(1:max(clusters$membership), function(i) {
  graph_subset <- keyword_cooccurrance[clusters$membership == i,clusters$membership == i]
  graph_subset <- graph.adjacency(graph_subset,
                                  weighted = TRUE,
                                  mode="undirected",
                                  diag=FALSE)
  eigens <- eigen_centrality(graph)$vector
  eigens_df <- tibble(eigenvalue = eigens, keyword = names(eigens), cluster = i) %>%
    arrange(-eigenvalue)
})

keyword_importance %>% group_by(cluster) %>%
  slice(1:10)

word_to_cluster <- tibble(keyword = V(graph)$name, cluster = clusters$membership)

assign_cluster <- tidy_tokens %>%
  filter(word %in% word_to_cluster$keyword) %>%
  inner_join(word_to_cluster, by = c("word" = "keyword")) %>%
  count(url_l, cluster, name = "votes") %>%
  group_by(url_l) %>%
  slice_max(votes, n = 1, with_ties = TRUE) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(cluster = factor(cluster, labels = paste("Cluster", sort(unique(cluster)))))

assign_cluster$cluster <- factor(assign_cluster$cluster,
                                 labels = c("Nature",
                                            "Wildlife",
                                            "Cold Weather",
                                            "Seasons",
                                            "People and the Park",
                                            "Birds",
                                            "Photos and Blogs",
                                            "Other"))

sutton_photo_clusters <- sutton_photos %>%
  inner_join(assign_cluster) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = st_crs(4326))
```

Flickr is an interesting data source for understanding people's
experience of green spaces. It provides a wide variety of tagged photos,
across a wide temporal and spatial range. These tags give an indicator
of what people associate with that particular photo. Furthermore,
location tags and timestamps give a precise point in space and time when
this occurred. However, it worth noting that the platform of Flickr is
mostly the realm of photographers, and is not as synonymous as a
platform such as Instagram.

```{r sutton photo cluster map, echo = F, message = F, warning = F, fig.width = 10, fig.cap = "Figure 10: A map showing the locations of photos of Sutton Park (Source: Flickr)"}
ggplot() +
  geom_sf(data = boundary) +
  geom_sf(data = sutton_photo_clusters, aes(colour = cluster))
```

There is an interesting spatial pattern for the cluster "People and the
Park", which clearly shows an outline of major paths in the park.
Furthermore, the "Wildlife" cluster is very centred around the SE of the
park. It is also worth noting that a shortfall of the machine learning
algorithm used to form these clusters may result in more positive or
negative clusters, potentially due to how it links similar words
together.

```{r sutton senitment analysis, echo = F, message = F, warning = F}
sutton_photo_clusters$text <- paste(sutton_photo_clusters$title,
                                    sutton_photo_clusters$description,
                                    sutton_photo_clusters$tags)

sutton_photo_text <- sutton_photo_clusters %>%
  unnest_tokens(word, text)

sentiment <- sutton_photo_text %>%
  inner_join(get_sentiments("afinn")) %>%
  group_by(url_l) %>%
  summarise(sentiment = sum(value)) %>%
  st_set_geometry(NULL) %>%
  inner_join(sutton_photo_clusters) %>%
  arrange(-sentiment) %>%
  mutate(id = 1:n())
```

```{r sutton senitment graph, echo = F, message = F, warning = F, fig.width = 10, fig.cap = "Figure 11: A graph to show sentiment analysis of the photo data of Sutton Park"}
ggplot(sentiment, aes(x = id, y = sentiment,
                      fill = cluster)) +
  geom_bar(stat = "identity")
```

To further this project, a data source such as Instagram may be useful
due its more widespread use, however their API is much more difficult to
access. Also of use may be a review platform, such as Google Maps.

## Final Project: Using GTFS Feeds to Visualise Transport Data from the West Midlands

**Project Aim**: Using Transport for West Midlands' GTFS feeds and the
tidytransit R package [@poletti2026], this project shows how service
frequency and transport mode varies across the West Midlands Combined
Authority area.

### Todo

1.  Briefly explain what you’re trying to achieve with the project

2.  Explain the choices you made on the techniques used

3.  Reflect on any ethical implications

4.  Critically evaluate the effectiveness of your project

## References
