---
title: "Digital Data Capture & Analysis Lab Book"
author: "2592586"
output: 
  html_document: 
    toc: true
    toc_depth: 6
    toc_float: true
    theme: paper
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r load packages, warning = F, message = F}
# general libraries to load
library(tidyverse)
library(sf)

# env. sensing libraries
library(plotly)
library(ggpubr)
library(ggpmisc)
library(cowplot)
library(lubridate)
library(colourpicker)
library(gganimate)
library(thermocouple)
library(Metrics)

# api + ai libraries
library(photosearcher)
library(textclean)
library(tidytext)
library(igraph, warn.conflicts = FALSE)

# final project libraries
library(tidytransit)
library(maptiles)
library(tidyterra)
```

## Creating a 360 Virtual Tour of Campus

This week we have been using [ThingLink](https://thinglink.com) and 360
photos taken with GoPro/Insta360 cameras to create 360 Virtual Tours.

Our tour is of the Y3 building on the University campus. We started in
the basement, going "up" the stairs and out of building, moving past the
front of the building and back in via another door. We then went back
into the basement via the lift, and then finishing in one of the seminar
rooms. We used a GoPro Max camera on a tripod, and connected it via
Bluetooth to an iPhone with the GoPro Quik app on for remote control of
the shutter.

I then uploaded the footage to ThingLink to create the tour. The
platform was fairly intuitive to use, but it was a slow and repetitive
task to go through each image in turn, changing settings and adding
"tags" to enable moving between each image. Furthermore, there were some
shots with people in, which were blurred out for privacy reasons.
Viewing it with a VR headset was impressive, despite the low resolution
of the images.

A virtual tour such as this would be useful for familiarising people
with (for example) a campus, office space, or a new development.
However, ethical concerns such as photographing of bystanders, as well
as taking photos of sensitive sites should be considered.

Below is the final tour on ThingLink.

```{r thinglink tour, echo = F, fig.width = 10, fig.cap = "Figure 1: A virtual tour of University of Birmingham Y3 Building"}
knitr::include_url("https://www.thinglink.com/view/scene/2073448908817695397")
```

## Photogrammetry

```{r sfm model, echo = F, fig.width = 10, fig.cap = "Figure 2: A model of an IKEA Marius stool"}
knitr::include_url("https://sketchfab.com/models/6c712d4d24a74af5b12fbec2aef62aae/embed")
```

The model above is of an IKEA Marius stool in my living room. Overall, I
would say this model is quite poor, as it does not adequately represent
the stool.

Firstly, the mesh of the model is quite poor, leading to holes and
floating geometry, as ghost points are formed. There was also erroneous
mesh points formed from carpet and wall shots. As well as this, the
texture of the model is very poorly captured, with none of the black
colour of the stool in the model.

Both of these issues could be due to the lack of a plain background
(making it more difficult for the software to distinguish what the focus
is), a poor photo distribution (i.e. the photos did not provide
sufficient coverage of the stool), or possibly poor camera quality. The
other point to consider is the software used - an open-source package
called [Regard3D](https://regard3d.org). Alternative methods of
processing could be considered, such as modifying parameters, however
considering the aforementioned background concerns, these would probably
be not as consequential in approving the final model output.

## Environmental Sensing

```{r environmental sensing - load data, warning = F, message = F}
# load data files
df_cal <- read_csv("Practical-Week7/NTC_cal_example.csv")
df_adc_water <- read_csv("Practical-Week7/ADC_water_bath_example.csv")
df_adc_air <- read_csv("Practical-Week7/ADC_air_logging_example.csv")
df_adc_refT <- read_csv("Practical-Week7/ADC_ref_temp.csv")

# Correct time formatting error in df_adc_refT data
df_adc_refT$Time <- parse_time(df_adc_refT$Time, "%H.%M.%S")

# Restructure the data
## ADC water bath
df_adc_water_L <- df_adc_water |>
  pivot_longer(!c(Time),
               names_to = "ADC_Bits", values_to = "Resistance")

df_adc_water_L <- df_adc_water_L |>
  mutate(min=minute(Time))

## ADC air
df_adc_air_L <- df_adc_water |>
  pivot_longer(!c(Time),
               names_to = "ADC_Bits", values_to = "Resistance")

df_adc_air_L <- df_adc_air_L |>
  mutate(min=minute(Time))
```

Turning to environmental sensors, we looked into how open-source
hardware and software and how their use has lowered the barriers to
sensing. This opens up possibilities to non-profits, for example, to use
sensors to back up their concerns with data. In our case, we used a
temperature sensor, composed of a thermistor attached to an Arduino
micro-computer (see Figure 1). The resistance across the thermistor
changes with temperature, and therefore if this resistance is measured a
temperature can be calculated from it.

![Figure 3: Temperature Sensor
Schematic](labbook_images/clipboard-3779242095.png)

\

### Calibration

The first part of the process involved calibration. Using four water
baths at different temperatures, the resistance read by the
Arduino-based sensor was recorded along with the temperature from a
reference probe, with ten readings taken from each bath. The results are
shown in Figure 2:

```{r resistance-temp graph, fig.width = 10, fig.cap = "Figure 4: A graph to show Resistance against Temperature as measured by an Arduino-based temperature sensor"}
ggplot(data = df_cal, mapping = aes(x = Resistance, y = Ref_temp)) + 
  geom_point(size=3.5, shape=1, colour="black") + 
  ylab("Water temperature (\u00b0C)") +
  xlab("Resistance (\u2126)")
```

We can see a clear pattern of an increase of resistance as temperature
decreases - this means that the thermistor used has a negative
temperature coefficient (an NTC thermistor).

Next, we will fit a calibration model to the data. The code below
creates a linear and non-linear (polynomial) model from the data.

```{r models}
# create models
linear_model <- lm(data = df_cal, 
                   formula = Ref_temp ~ Resistance)

poly_model <- lm(data = df_cal, 
                 formula = Ref_temp ~ poly(
                   x = Resistance, degree = 2, raw = T))
```

The summaries are shown in the table below:

| Model | Residual Std. Error | Multiple R-Squared | F-Statistic and p-value |
|:--:|:--:|:--:|:--:|
| Linear Model | 3.784 (28 *df*) | 0.9434 | 466.4, p \< 2.2e-16 |
| Polynomial Model | 0.2554 (27 *df*) | 0.9998 | 5.424 × 10^4^ , p \< 2.2e-16 |

: Table 1: Calibration Model Summaries

The summary statistics from this model shows that the polynomial model
is a better fit to the data, with a smaller residual standard error, and
a larger R-Squared value and F-Statistic.

Below are the models plotted against the data:

```{r linear model and data graph, warning = F, message = F, fig.width = 10, fig.cap = "Figure 5: A graph to show Resistance against Temperature with a linear model"}
ggplot(df_cal, aes(y = Ref_temp, x = Resistance)) +
  geom_point() +
  geom_smooth(method = lm)
```

```{r polynomial model and data graph, warning = F, message = F, fig.width = 10, fig.cap = "Figure 6: A graph to show Resistance against Temperature with a non-linear model"}
ggplot(df_cal, aes(y = Ref_temp, x = Resistance)) +
  geom_point() +
  geom_smooth(method = lm, formula = y ~ poly(x, 2, raw = T))
```

The mean and standard deviation for each model is shown below.

```{r beta and log models; fitted values into a table, warning = F, message = F}
# create beta and log models
df_cal$Log_res <- log(df_cal$Resistance)
log_linear_model<-lm(data = df_cal, formula = Ref_temp ~ Log_res)
df_cal$Temp_beta <- ThermistorTemperature(R0 = 10000, T0 = 25, R = df_cal$Resistance, betaTH = 3950)
df_cal$Temp_linear <- linear_model$fitted.values
df_cal$Temp_poly <- poly_model$fitted.values
df_cal$Temp_loglin <- log_linear_model$fitted.values

# model summary table
tib_cal <- tibble(Ref_temp = df_cal$Ref_temp, Temp_beta = df_cal$Temp_beta, Temp_linear = df_cal$Temp_linear, Temp_poly = df_cal$Temp_poly, Temp_loglin = df_cal$Temp_loglin)

tib_cal_summary <- tib_cal |>
  group_by(Ref_temp) |>
  summarise(
    Mean_temp_beta = mean(Temp_beta),
    Sd_temp_beta = sd(Temp_beta),
    Mean_temp_linear = mean(Temp_linear),
    Sd_temp_linear = sd(Temp_linear),
    Mean_temp_poly= mean(Temp_poly),
    Sd_temp_poly = sd(Temp_poly),
    Mean_temp_loglin = mean(Temp_loglin),
    Sd_temp_loglin = sd(Temp_loglin)
  )

tib_cal_summary <- tib_cal_summary |> mutate(
  across(where(is.numeric), round, 3)
)

tib_cal_summary <- tib_cal_summary |>
  unite(Summ_temp_beta, Mean_temp_beta:Sd_temp_beta, sep = "±", na.rm = T) |>
  unite(Summ_temp_linear, Mean_temp_linear:Sd_temp_linear, sep = "±", na.rm = T) |>
  unite(Summ_temp_poly, Mean_temp_poly:Sd_temp_poly, sep = "±", na.rm = T) |>
  unite(Summ_temp_loglin, Mean_temp_loglin:Sd_temp_loglin, sep = "±", na.rm = T)

knitr::kable(tib_cal_summary, col.names = c("Ref Temp (°C)", "Beta Model (Mean ± SD °C)", "Linear Model (Mean ± SD °C)", "Poly Model (Mean ± SD °C)", "Log Linear Model (Mean ± SD °C)"), caption = "Table 2: Summary Statistics for each Reference Temperature and Model")
```

```{r calculate values, echo = T, warning = F, message = F, results = "hide", fig.width = 10, fig.cap = "Figure 7: Models Comparison"}
# model comparisons plot and stats
df_cal_long <- df_cal |> pivot_longer(!c(Cal_medium, Resistance,
                                       Voltage, Ref_temp, 
                                       Log_res),
                                    names_to = "cal_model", 
                                    values_to = "cal_temp")
beta_model <- ThermistorTemperature(R0 = 10000, T0 = 25, R = df_cal$Resistance, betaTH = 3950)

ggplot(data = df_cal_long,
       mapping = aes(x=Ref_temp, y = cal_temp,
       colour = cal_model))+geom_point(size=2)+
  facet_wrap(~cal_model,scales = "fixed")+
  theme_pubclean()+
  geom_abline(slope = 1)+
  geom_smooth(se = F,method = "lm",linetype = "dashed")+  
  stat_regline_equation(label.x=5, label.y=40)

summary(beta_model)
summary(linear_model)
summary(poly_model)
summary(log_linear_model)

rmse(df_cal$Ref_temp, df_cal$Temp_beta)
rmse(df_cal$Ref_temp, df_cal$Temp_linear)
rmse(df_cal$Ref_temp, df_cal$Temp_poly)
rmse(df_cal$Ref_temp, df_cal$Temp_loglin)
```

| **Variable** | Beta Model | Linear Model | Poly Model | Log Linear Model |
|--------------|------------|--------------|------------|------------------|
| R^2^         | N/A        | 0.9434       | 0.9998     | 0.9983           |
| RMSE         | 2.253915   | 3.655814     | 0.2423138  | 0.6319984        |
| Slope        | 0.86       | 0.94         | 1          | 1                |

: Table 3: Statistical Variables for Each Model

Table 3 shows that the sensor is quite accurate in calculating
temperature, especially when paired with a polynomial model. Whilst this
may not be as accurate as a commercial sensor, it gets incredibly close
and for most projects this should be fine.

### Analogue-Digital Conversion (ADC)

```{r adc boxplots data, warning = F, message = F}
# Note: I used the Water data for this section as opposed to the Air data, since the joining of the air temperature and air reference temperature datasets failed, as there were no common times between them.

# look at adc data and models
df_adc_water_L$Temp <- predict(object = poly_model, df_adc_water_L)

df_adc_air_L$Temp <- predict(object =poly_model, df_adc_air_L)

df_adc_air_L$Temp_beta <-ThermistorTemperature(R0 = 10000, T0 = 25, R = df_adc_air_L$Resistance, betaTH = 3950)

df_adc_water_L$Temp_beta <-ThermistorTemperature(R0 = 10000, T0 = 25, R = df_adc_water_L$Resistance, betaTH = 3950)

df_adc_ref_water <- df_adc_refT |> 
  filter(Medium == "Water") |> 
  left_join(df_adc_water_L, by = "Time")

df_adc_ref_air <- df_adc_refT |> 
  filter(Medium == "Air") |> 
  left_join(df_adc_air_L, by = "Time")
```

```{r adc boxplot one, warning = F, message = F, fig.width = 10, fig.cap = "Figure 8: Plot of ADC (using Water reference)"}
df_adc_ref_water |> 
  filter(!is.na(ADC_Bits)) |>
  ggplot(aes(x = ADC_Bits, y = Temp_beta)) + 
  geom_boxplot() + 
  geom_point()
```

Figure 8 shows that the 8-bit ADC conversion returns a slightly higher
value than a 10-bit conversion, and 12-bit is lower than 10-bit.
However, the difference between the 10-bit and 12-bit conversions is
much less than that between 8-bit and 10-bit conversions. This is shown
more clearly in Figure 9 below.

```{r adc boxplot two, warning = F, message = F, fig.width = 10, fig.cap = "Figure 9: Plot of ADC (Using Water reference, without box)"}
df_adc_ref_water |> 
  filter(!is.na(ADC_Bits)) |>
  ggplot(aes(x = ADC_Bits, y = Temp_beta)) + 
  geom_dotplot(binaxis = "y")
```

```{r adc summary table, warning = F, message = F}
# summary table for adc data
tib_adc_ref_water <- tibble(df_adc_ref_water)

tib_adc_summary <- tib_adc_ref_water |>
  filter(!is.na(ADC_Bits)) |> 
  group_by(ADC_Bits) |>
  summarise(
    Mean = mean(Temp_ref),
    Sd = sd(Temp_ref, na.rm = T)
  )

tib_adc_summary <- tib_adc_summary |> mutate(
  across(where(is.numeric), round, 3)
)

tib_adc_summary <- tib_adc_summary |>
  unite(Summ_temp_ref, Mean:Sd, sep = "±", na.rm = T)

knitr::kable(tib_adc_summary, caption = "Table 5: Summary Statistics for each Resolution")
```

It is interesting to note that the summary statistics for each
resolution are the same, suggesting that the difference is not noted
when rounded to 3 decimal places.

Looking at the sensor setup overall, the Arduino platform is a
well-developed "open-hardware" ecosystem with supporting software and
documentation. However, as with all "open-source" systems, there are
some pitfalls. Firstly, the system is very much DIY - you are provided
with the components, but as the user you must then assemble the parts,
both physically (the hardware) and in terms of programming the assembled
system to do what you would like it to. This requires, in the case of
the Arduino ecosystem, a knowledge of C/C++ languages. Furthermore, if
anything goes wrong, there is no dedicated customer support, but only
documentation and support forums. That being said, if these obstacles
can be overcome, the Arduino platform can scale rapidly and relatively
cheaply compared to alternative proprietary sensing systems, making it a
viable option for more widespread sensing networks, or for smaller
companies and/or non-profits who cannot afford the more expensive
alternatives.

## APIs and AI - Using Flickr's API to map sentiments and topics spatially in Sutton Park

Looking at how to use APIs and AI, this next section shows how the use
of [Flickr](https://flickr.com)'s API can show in a visual way how
people experience green spaces, with very precise spatial data.

```{r apis and ai setup, message = F, warning = F, results = "hide"}
boundary <- st_read("Practical-Week6/Sutton Park Boundary_region.shp")

# note - I saved the data from the Flickr API to a csv file in order to save rendering time, as fetching all the data from the API every time takes a while!
sutton_photos <- read_csv("Practical-Week6/sutton_photos.csv", col_names = T, na = "NA")

sutton_text <- sutton_photos |>
  mutate(
    text = paste(title, description, tags),
    text = replace_url(text),
    text = str_squish(text)
  ) |>
  filter(!is.na(text), text != "")

data("stop_words")

tidy_tokens <- sutton_text |>
  select(url_l, longitude, latitude, title, description, tags, text) |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  filter(str_detect(word, "^[a-z][a-z'-]*$")) |>
  count(url_l, word, sort = FALSE)

photo_word <- tidy_tokens |>
  mutate(n = 1) |>
  pivot_wider(names_from = word, values_from = n, values_fill = 0)

word_counts <- tidy_tokens |> count(word, name = "photo_count")

keep_words <- word_counts |> filter(photo_count >= 5) |> pull(word)

photo_word_mat <- tidy_tokens |>
  filter(word %in% keep_words) |>
  mutate(n = 1) |>
  pivot_wider(names_from = word, values_from = n, values_fill = 0) |>
  select(-url_l) |>
  as.matrix()

keyword_cooccurrance <- t(photo_word_mat) %*% photo_word_mat

graph <- graph_from_adjacency_matrix(keyword_cooccurrance,
                         weighted = TRUE,
                         mode="undirected",
                         diag=FALSE)

clusters <- cluster_walktrap(graph)

keyword_importance <- map_df(1:max(clusters$membership), function(i) {
  graph_subset <- keyword_cooccurrance[clusters$membership == i,clusters$membership == i]
  graph_subset <- graph.adjacency(graph_subset,
                                  weighted = TRUE,
                                  mode="undirected",
                                  diag=FALSE)
  eigens <- eigen_centrality(graph)$vector
  eigens_df <- tibble(eigenvalue = eigens, keyword = names(eigens), cluster = i) |>
    arrange(-eigenvalue)
})

keyword_importance |> group_by(cluster) |>
  slice(1:10)

word_to_cluster <- tibble(keyword = V(graph)$name, cluster = clusters$membership)

assign_cluster <- tidy_tokens |>
  filter(word %in% word_to_cluster$keyword) |>
  inner_join(word_to_cluster, by = c("word" = "keyword")) |>
  count(url_l, cluster, name = "votes") |>
  group_by(url_l) |>
  slice_max(votes, n = 1, with_ties = TRUE) |>
  slice(1) |>
  ungroup() |>
  mutate(cluster = factor(cluster, labels = paste("Cluster", sort(unique(cluster)))))

assign_cluster$cluster <- factor(assign_cluster$cluster,
                                 labels = c("Nature",
                                            "Wildlife",
                                            "Cold Weather",
                                            "Seasons",
                                            "People and the Park",
                                            "Birds",
                                            "Photos and Blogs",
                                            "Other"))

sutton_photo_clusters <- sutton_photos |>
  inner_join(assign_cluster) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = st_crs(4326))
```

Flickr is an interesting data source for understanding people's
experience of green spaces. It provides a wide variety of tagged photos,
across a wide temporal and spatial range. These tags give an indicator
of what people associate with that particular photo. Furthermore,
location tags and timestamps give a precise point in space and time when
this occurred. However, it worth noting that the platform of Flickr is
mostly the realm of photographers, and is not as synonymous as a
platform such as Instagram.

```{r sutton photo cluster map, message = F, warning = F, fig.width = 10, fig.cap = "Figure 10: A map showing the locations of photos of Sutton Park (Source: Flickr)"}
ggplot() +
  geom_sf(data = boundary) +
  geom_sf(data = sutton_photo_clusters, aes(colour = cluster)) + 
  labs(colour = "Cluster") + 
  theme_bw()
```

There is an interesting spatial pattern for the cluster "People and the
Park", which clearly shows an outline of major paths in the park.
Furthermore, the "Wildlife" cluster is very centred around the SE of the
park. It is also worth noting that a shortfall of the machine learning
algorithm used to form these clusters may result in more positive or
negative clusters, potentially due to how it links similar words
together.

```{r sutton senitment analysis, message = F, warning = F}
sutton_photo_clusters$text <- paste(sutton_photo_clusters$title,
                                    sutton_photo_clusters$description,
                                    sutton_photo_clusters$tags)

sutton_photo_text <- sutton_photo_clusters |>
  unnest_tokens(word, text)

sentiment <- sutton_photo_text |>
  inner_join(get_sentiments("afinn")) |>
  group_by(url_l) |>
  summarise(sentiment = sum(value)) |>
  st_set_geometry(NULL) |>
  inner_join(sutton_photo_clusters) |>
  arrange(-sentiment) |>
  mutate(id = 1:n())
```

```{r sutton senitment graph, message = F, warning = F, fig.width = 10, fig.cap = "Figure 11: A graph to show sentiment analysis of the photo data of Sutton Park"}
ggplot(sentiment, aes(x = id, y = sentiment,
                      fill = cluster)) +
  geom_bar(stat = "identity") + 
  labs(x = "ID", y = "Sentiment", fill = "Cluster") + 
  theme_bw()
```

To further this project, a data source such as Instagram may be useful
due its more widespread use, however their API is much more difficult to
access. Also of use may be a review platform, such as Google Maps.

## Final Project: Using GTFS Feeds to Visualise Transport Data from the West Midlands

```{r tfwm project data, message = F, warning = F, results = "hide"}
# load gtfs data from zip downloaded from tfwm api
tfwm <- read_gtfs("Final-Project/tfwm.zip")
gtfs <- set_servicepattern(tfwm)
gtfs <- gtfs_as_sf(gtfs)

gtfs$shapes$length <- st_length(gtfs$shapes)

shape_lengths <- gtfs$shapes |> 
  as.data.frame() |> 
  select(shape_id, length, -geometry)

# create service pattern summary data frame
service_pattern_summary <- gtfs$trips |> 
  left_join(gtfs$.$servicepatterns, by="service_id") |> 
  left_join(shape_lengths, by="shape_id") |>
  left_join(gtfs$stop_times, by="trip_id") |> 
  group_by(servicepattern_id) |> 
  summarise(
    trips = n(), 
    routes = n_distinct(route_id),
    total_distance_per_day_km = sum(as.numeric(length), na.rm=TRUE)/1e3,
    route_avg_distance_km = (sum(as.numeric(length), na.rm=TRUE)/1e3)/(trips*routes),
    stops=(n_distinct(stop_id)/2))

service_pattern_summary <- gtfs$.$dates_servicepatterns |>  
  group_by(servicepattern_id) |> 
  summarise(days_in_service = n()) |> 
  left_join(service_pattern_summary, by="servicepattern_id")

service_ids <- gtfs$.$servicepatterns |>
  pull(service_id)

gtfs$trips |>
  filter(service_id %in% service_ids) |>
  group_by(service_id, route_id) |>
  left_join(gtfs$routes, by = "route_id") |>
  filter(agency_id == "OP100") |>
  summarise(count = n())

am_stop_freq <- get_stop_frequency(gtfs, start_time = 6*3600, end_time = 10*3600, 
                                   service_ids = service_ids, by_route = TRUE)
```

**Project Aim**: Using [Transport for West Midlands (TfWM)'s
API](https://api-portal.tfwm.org.uk) and the *tidytransit* R package
[@poletti2026], this project aims to visualise key and useful aspects of
the [National Express Bus West Midlands (NXBus WM)
network](https://nxbus.co.uk/west-midlands).

The General Transit Feed Specification (GTFS) is a specification that
enables transport providers to provide service information in a uniform
manner, to be accessible to third-party applications through an API,
enabling accurate service information in apps such as Google Maps. The
*tidytransit* package enables R to interact with the GTFS format,
enabling it to be used in data analysis and visualisation. This project
is based on some of the [*tidytransit* sample
projects](https://r-transit.github.io/tidytransit/articles/introduction.html),
although I have modified a substantial portion to create interesting
visualisations for this particular dataset.

Firstly - due to the number of routes run by NXBus WM and destinations
covered - I have filtered the dataset to only include their express "X"
services. These routes are shown in Figure 12.

```{r nxbus routes + data prep, warning = F, message = F, results = "hide", fig.width = 10, fig.cap = "Figure 12: Express Routes"}
# filter for stops on lines
one_line_stops <- am_stop_freq |>
  left_join(gtfs$routes, by = "route_id") |>
  filter(agency_id == "OP100", direction_id == 0) |>
  filter(str_detect(route_short_name, "X")) |>
  left_join(gtfs$stops, by ="stop_id") |> 
  mutate(mean_headway_minutes = mean_headway/60)

# get spatial data for routes
one_line_stops_sf <- gtfs$stops |>
  right_join(one_line_stops, by="stop_id")

# convert to correct crs
one_line_stops_sf_crs <- sf::st_transform(one_line_stops_sf, 4326)

# get online tile for basemap
basemap <- get_tiles(one_line_stops_sf_crs, provider = "CartoDB.Positron", zoom = 12)

# get frequency of routes
am_route_freq <- get_route_frequency(gtfs, service_ids = service_ids, 
                                     start_time = 6*3600, end_time = 10*3600)

# get spatial data for routes
routes_sf <- get_route_geometry(gtfs, service_ids = service_ids)
routes_sf <- routes_sf |> 
  inner_join(am_route_freq, by = "route_id")

# convert to correct crs
routes_sf_crs <- sf::st_transform(routes_sf, 4326)

nxbus_express <- tfwm$routes |>
  filter(agency_id == "OP100") |>
  filter(str_detect(route_short_name, "X"))

routes_sf_crs <- routes_sf_crs |>
  inner_join(nxbus_express, by = "route_id")

routes_sf_crs |>
  ggplot() + 
  geom_spatraster_rgb(data = basemap) + 
  geom_sf(aes(colour = route_short_name)) + 
  labs(color = "Route", caption = "© OpenStreetMap contributors © CARTO © Transport for West Midlands") +
  theme_bw()
```

Next, Figure 13 shows a map of the mean headway for each stop on the
line. Headway is the distance between vehicles on the line. It is
interesting to note that the headway decreases the closer to the centre
of Birmingham the lines go.

```{r nxbus stops and headways, warning = F, message = F, results = "hide", fig.width = 10, fig.cap = "Figure 13: Mean Headway Times at Stops"}
ggplot() + 
geom_spatraster_rgb(data = basemap) +
geom_sf(data = one_line_stops_sf_crs, aes(color = mean_headway_minutes)) +
labs(color = "Mean Headway (Minutes)", caption = "© OpenStreetMap contributors © CARTO © Transport for West Midlands") + 
theme_bw()
```

Turning to departure frequency across the stops on the express services
(between 09:00 and 17:00), this visualisation shows the end locations of
services throughout the day, including how service termini changes
across the day.

```{r nxbus departure frequencies, message = F, warning = F, fig.width = 10, fig.cap = "Figure 14: Destinations and Frequency"}
# get data for nxbus routes and stops
nxbus <- tfwm$routes |>
  filter(agency_id == "OP100")

nxbus_trips <- left_join(tfwm$trips, nxbus, by="route_id") |>
  filter(agency_id == "OP100")

nxbus_stoptimes <- left_join(tfwm$stop_times, nxbus_trips, by="trip_id") |>
  filter(agency_id == "OP100")

nxbus_stops <- left_join(tfwm$stops, nxbus_stoptimes, by="stop_id") |>
  filter(agency_id == "OP100")

nxbus_stops_time_sep <- nxbus_stops |>
  separate(departure_time, sep = ":", into = c("dep_hours", "dep_mins", "dep_secs")) |>
  mutate_at(c("dep_hours", "dep_mins", "dep_secs"), as.numeric)

# filter by departures between 09:00 and 17:00
departures_nine_to_five <- nxbus_stops_time_sep |>
  filter(dep_hours >= 9 & dep_hours <= 16)

graph_data <- departures_nine_to_five |>
  mutate(departure_time = make_datetime(hour = dep_hours, min = dep_mins, sec = dep_secs)) |>
  filter(str_detect(route_short_name, "X"))

ggplot(graph_data) + theme_bw() +
  geom_point(aes(y=trip_headsign, x=departure_time, color = route_short_name), size = 0.2) +
  scale_x_time(breaks = seq(0, max(as.numeric(graph_data$departure_time)), 3600), 
               labels = scales::time_format("%H")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position = "bottom") + 
  guides(color = guide_legend(override.aes = list(size = 3))) + 
  labs(x = "Departure Time", y = "Destination", color = "Route", caption = "© Transport for West Midlands")
```

This data can be further filtered to look at departures from a
particular stop. A common stop for buses is Birmingham New Street
station: departures from here, on 1st April, are shown below.

```{r departures from new st station, warning = F, message = F, fig.width = 10, fig.cap = "Figure 15: Departures from Birmingham New St Station (01-04-26)"}
# add trip origin data to dataset
first_stop_id <- gtfs$stop_times %>% 
  group_by(trip_id) %>% 
  summarise(stop_id = stop_id[which.min(stop_sequence)])

first_stop_names <- left_join(first_stop_id, gtfs$stops, by="stop_id")

trip_origins <- first_stop_names %>% select(trip_id, trip_origin = stop_name)

gtfs$trips <- left_join(gtfs$trips, trip_origins, by = "trip_id")

# filter for stop New Street
stop_ids <- gtfs$stops |> 
  filter(stop_name == "New Street" | stop_name == "New St Station" | stop_name == "New St") |> 
  select(stop_id)

# create departures data frame
departures <- stop_ids |> 
  inner_join(gtfs$stop_times |> 
               select(trip_id, arrival_time, 
                      departure_time, stop_id), 
             by = "stop_id")

departures <- departures |> 
  left_join(gtfs$trips |> 
              select(trip_id, route_id, 
                     service_id, trip_headsign, 
                     trip_origin), 
            by = "trip_id")

departures <- departures |> 
  left_join(gtfs$routes |> 
              select(route_id, 
                     route_short_name, agency_id), 
            by = "route_id")

# filter to services on 01-04-26
services_on_010426 <- gtfs$.$dates_services |> 
  filter(date == "2026-04-01") |> 
  select(service_id)

# filter departures by services on 01-04-26
departures_010426 <- departures |> 
  inner_join(services_on_010426, by = "service_id") |>
  filter(agency_id == "OP100") |>
  filter(str_detect(route_short_name, "X"))

ggplot(departures_010426) +
  theme_bw() + 
  geom_point(aes(y = route_short_name, x = departure_time, color = route_short_name), size = 2) + 
  scale_x_time(breaks = seq(0, max(as.numeric(departures_010426$departure_time)), 3600), labels = scales::time_format("%H")) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = "Departure Time", y = "Route", caption = "© Transport for West Midlands") + 
  theme(legend.position = "none")
```

These visualisations help to show the distribution of buses across
Birmingham, and frequency data for buses. This data could be useful to
transport planners in looking at the different bus services in the
region. However, a limitation is that TfWM does not provide a live GTFS
feed as part of their public API, and therefore data such as punctuality
and reliability cannot be analysed as part of this project.

## References
